The five pillars of computational reproducibility: Bioinformatics and beyond

### 2023-08-15 manuscript/rebuttal.Rmd

> Comments to the Author: The article offers a comprehensive perspective on reproducibility in
> computational research, delivering clear and logically backed recommendations. It is indeed a very
> interesting topic that needs our attention. Nevertheless, there are areas that could benefit from
> further development and clarity:

Response: We appreciate reviewer 2's positive feedback, insights and thoughtful suggestions.

> 1. Inclusion of Case Studies or Examples: To strengthen your argument, consider incorporating case
> studies or practical examples that demonstrate the effectiveness of your recommendations.
> Highlighting specific instances where these best practices have enhanced reproducibility or solved
> particular issues could substantiate your points more effectively.

To demonstrate the importance of computational reprodcibility, we elaborate on the
"forensic bioinformatics" case study of Baggerly and Coombes (2010) critique of Potti et al (2006)
in the introduction:

@@

> 2. Expanding on Current Shortcomings: The article mentions that current practices are not meeting the
> goal of reproducibility. An expanded discussion on the ramifications of these shortcomings on
> scientific progress could accentuate the urgency and significance of addressing this issue.

In the introduction we have spoken to the consequences of poor computational research, and emphasize
the urgency of this issue (which complements the case study above):

> The ramifications of irreproducible and unreliable research include misleading the community,
> wasting research funds, slowing scientific progress, eroding public confidence in science and
> tarnishing the reputation of associated institutions and colleagues.
> In certain cases, irreproducible bioinformatics has the potential to place patient safety at risk.

### 2023-08-15 manuscript/5pillars.Rmd
challenges and strategies

Due to the constantly developing software landscape, this review is not intended to give strong
recommendations on the exact combination of tools or resources to use in computational research,
but to set out key principles and highlight some useful solutions that users can incorporate
into their toolkit to improve the overall reproducibility of their work.

@@

The key technological advances enabling the five pillars approach have been around for about a decade,
yet progress on integrating these into computational research publications has been minimal.
Indeed the number of systematic replication studies in bioinformatics remains small
[@Ioannidis2009-hv; @Zaringhalam2020-ms; @Samuel2022-ha], so the severity of the problem isn't
well understood.
Part of the problem is the incentive structures in academia, with funding bodies and journals
largely focused on novel research directions rather than confirming previous findings
[@Garijo2013-zl ; @Huang2013-xr; @Kaelin2017-id; @Munafo2017-ty; @Diong2021-is].

(TODO: reviewer 2 point 3: focus on implementation challenges and strategies to overcome.
Expand this section articulating the difficulties with making gold standard reproducibility the norm.
Detail some strategies that can be used.
Case in point: spreadsheet causing gene name errors

1. described in 2004

2. followed by large scale screens in 2016,2022

3. HUGO finally takes action

4. Public awareness

5. Numbers of affected articles diminishes
)

A lack of technical computer skills in many research groups is another culprit.
Specific workshop-based training in these areas could help [@Williams2023-ic].
The [Software Carpentry](https://software-carpentry.org/) initiative is focused on this mission
[@Wilson2014-mh].
However, the most scalable solution would be to provide comprehensive beginner-friendly step-by-step
guides for frequently used routines.
For example, our group has just released such a protocol that walks novice users through the process
of implementing functional enrichment analysis of gene expression data with the five pillars principles
@Ziemann2023-nm.
More such protocols are needed to make extreme reproducibility mainstream.
A compendium of publicly available learning materials around the five pillar concepts are
provided in the supplementary information.

>>>>

Progress on implementing these best practices into computational research publications has been minimal, as
researchers face a range of obstacles.

These include:

* The "publish-or-perish" incentive structures in academia reward prolific rather than scrupulous
authors.
Part of the problem is the incentive structures in academia, with funding bodies and journals
largely focused on novel research directions rather than confirming previous findings
[@Garijo2013-zl ; @Huang2013-xr; @Kaelin2017-id; @Munafo2017-ty; @Diong2021-is].

* Journal standards are slow to progress.
To illustrate this, many journals still allow publication without deposition of code or data to public
repositories.

* A lack of systematic reproducibility/replication studies in bioinformatics remains small
[@Ioannidis2009-hv; @Zaringhalam2020-ms; @Samuel2022-ha], so the severity of the problem isn't
well understood.

* A lack of documentated protocols and tutorial materials.

* An overall lack of coding/data science expertise among life science researchers and new graduates.

Overcoming these challenges will require a multi-pronged strategy:

* Incentive structures will need to change.
How we measure and value research achievement.
A change in culture or research assessment.
How we fund research data analytics and infrastructure.
Value rigour as well as prolificness.

* Journals need to progress towards publishing reproducible workflows.
Peng 2011 speaks about the lack of infrastructure for disseminating reproducible research.

* Enhanced training.
A lack of technical computer skills in many research groups is another culprit.
Specific workshop-based training in these areas could help [@Williams2023-ic].
The [Software Carpentry](https://software-carpentry.org/) initiative is focused on this mission
[@Wilson2014-mh].
Undergraduate and postgraduate courses in the life sciences would benefit from enhanced training in data
science.

* Enhanced documentation, protocols and tutorials.
The most scalable solution would be to provide comprehensive beginner-friendly step-by-step
guides for frequently used routines.
For example, our group has just released such a protocol that walks novice users through the process
of implementing functional enrichment analysis of gene expression data with the five pillars principles
@Ziemann2023-nm.
More such protocols are needed to make extreme reproducibility mainstream.
A compendium of publicly available learning materials around the five pillar concepts are
provided in the supplementary information.

* What individuals can do.

* Overall change in culture.
Recognition that life sciences is a data intensive discipline.

## Future directions

@@
